Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model

### 2022-08-03 Rmarkdown/paper.Rmd
edits from coauthors

$$(1)~~y = \beta_{0} + f(x_t) + \epsilon_t$$ 
where $f(x_t)$ is a smooth function of $x_t$,

$$(2)~~f(x_t) = \sum_{j=1}^kb_j(x_t)\beta_j$$
$b_j(x_t)$ are the functions of $x_t$ on which $y$ is regressed, $\beta_j$ are the model coefficients, and $\epsilon_t$ is the error term. We fitted the example GAM in Figures 2--3 using the R package *mgcv* [@mgcv], which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.

@@

<!-- $$\begin{align} -->
<!-- P(\theta|D) \propto P(D|\theta)P(\theta) \\ -->
<!-- posterior \propto likelihood \times prior \\ -->
<!-- posterior = P(\theta | D) \\ -->
<!-- likelihood = P(D | \theta) \\ -->
<!-- prior = P(\theta) -->
<!-- \end{align}$$ -->

@@

$$
(5)~~
\begin{align}
likelihood: \\
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu) \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu) \\
y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
\\
model: \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
priors: \\
\sigma \sim T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

@@

Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $x$, *at most*) [@stan2021; @mcelreath_statistical_2016]. Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively. Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
\alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
\sigma_{\alpha} \sim T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:

$$
(7)~~
\begin{align}
generalized~additive~model~(GAM): \\
f_j(t) = X_j\beta_j + Z_jb_j \\
\\
priors~for~GAM~coefficients: \\
\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
b_j \sim N(0, \sigma_b)~~[penalized] \\
\sigma_b \sim N(0, 0.5) (seasonal~smoother) \\
\sigma_b \sim T(0, 2.5, 3) (global,~group-level~smoothers) \\ 
\end{align}
$$

@@

The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception. This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].

@@

@@

Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$. The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but still reflected in model predictions: see Figure 6 and equation 5).

@@

Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6). During the conditioning period in particular---before orthophosphate was introduced---particles represented a minor contribution. Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020]. Given the prominent seasonality in dissolved lead release, the particulate fraction was inversely seasonal, with higher particulate fractions in winter (Figure 6). The increase in particulate lead in the final weeks of the study is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].

Effluent lead concentrations in pipe loop studies often decrease sharply after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to nonseasonal changes in water quality (see *Trends in particulate lead release*).

The variation in dissolved lead captured in Figure 8a neglects seasonality by design. And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016]. And although dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).

@@

Our approach can be modified to suit the needs of diverse analyses. While time is the sole independent variable here, water quality parameters can easily be added as predictors if inferences concerning their effects are of interest. Moreover, any parameter in the likelihood can be modeled---including the standard deviation if model residuals are heteroskedastic. The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical]. But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time applications.

>>>>

$$(1)~~y_t = \beta_{0} + f(t) + \epsilon_t$$ 
where $\beta_0$ is the intercept, $f(t)$ is a smooth function of $t = time$,

$$(2)~~f(t) = \sum_{j=1}^kb_j(t)\beta_j$$
$b_j(t)$ are the basis functions of $t$ that comprise $f(t)$, $\beta_j$ are the model coefficients (i.e., the basis function weights), and $\epsilon_t$ is the error term. We fitted the example GAM in Figures 2--3 using the R package *mgcv* [@mgcv], which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.

@@

@@

$$
(5)~~
\begin{align}
likelihood: \\
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu)~~[observed] \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu)~~[left \text{-} censored] \\
y_t | censored_t = 2 \sim g(\mu_t,\sigma,\nu)~~[interval \text{-} censored] \\
\\
model: \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
priors: \\
\sigma \sim Half \text{-} T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

@@

Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq l)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $l$, *at most*) [@stan2021; @mcelreath_statistical_2016]. Interval-censored observations are indicated by $censored = 2$ and were modeled using $g(l, u)$, or $P(l \lt X \leq u) = F_X(u) - F_X(l)$, where $l$ and $u$ are the lower and upper limits, respectively. Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(l)$, only.

@@

$$
(6)~~
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
\alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
\end{align}
$$
The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$. The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:

$$
(7)~~
\begin{align}
generalized~additive~model~(GAM): \\
f_j(t) = X_j\beta_j + Z_jb_j \\
\\
priors~for~GAM~coefficients: \\
\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
b_j \sim N(0, \sigma_b)~~[penalized] \\
\sigma_b \sim Half \text{-} N(0, 0.5) (seasonal~smoother) \\
\sigma_b \sim  \text{-} T(0, 2.5, 3) (global/group \text{-}level~smoothers) \\ 
\end{align}
$$

@@

The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5^th^ percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception. This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].

@@

In a typical pipe loop study, effluent lead concentrations decrease sharply after commissioning. Assuming inputs to the system remain constant, a period of relative stability usually follows [@wysock_statistical_1995]. This pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured by the model (see *Trends in particulate lead release*).

The variation in dissolved lead evident in Figure 8a ignores seasonality by design. And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016]. And although dissolved organic carbon has also been shown to increase lead solubility [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).

Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6). During the conditioning phase in particular---before orthophosphate was introduced---particles represented a minor contribution. Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020]. The prominent seasonality in dissolved lead release implies an inversely seasonal particulate fraction (Figure 6). That is, particles contributed more to the total lead concentration in winter.

The decrease and subsequent increase in particulate lead beginning in late 2020 are not easily explained, but they may be related to water temperature. Seasonal temperature variation was most pronounced in the final 1.5 years of the study (Figure 8c), coinciding approximately with the apparent seasonal pattern in particulate lead (Figure 8a). Increasing feedwater aluminum concentrations may also have been a factor (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].

@@

Autocorrelation models the dependence of observations on previous observations, or the response of a measurement series at time $t+s$ to a shock at time $t$. The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but still reflected in model predictions: see Figure 6 and equation 5).

@@

@@

Our approach can be modified to suit the needs of diverse analyses. While time is the sole independent variable here, water quality parameters can easily be added as predictors if inferences concerning their effects are of interest. Moreover, any parameter in the likelihood can be modeled---including the standard deviation if model residuals are heteroskedastic. The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical]. But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time.
